\section*{Data Availability}
All datasets reanalysed in this study are publicly available through ProteoMeXchange.
Cell-line bulk DIA reanalyses: PRIDE PXD003539 (Guo 2019, NCI-60), PXD004701 (Sun 2023), PXD030304 (ProCan-DepMapSanger), PXD017199 (Tognetti 2021), PXD041421 (Wang 2023).
Scalability experiment: PRIDE PXD071075.
ProteoBench benchmarks: PXD049412 (single-cell, Module 9), PXD062685 (diaPASEF, Module 5), PXD070049 (ZenoTOF, Module 10), and Module 7 (Astral) datasets.
The full list of PRIDE accessions used in the pan-cohort of DIA cell-line reanalyses is provided in Additional file~1: Table~S1.
\quantmsdiann{} is available at \url{https://github.com/bigbio/quantmsdiann} under the MIT license; documentation is hosted at \url{https://quantmsdiann.quantms.org/}.
The pipeline is implemented in Nextflow DSL2 ($\geq$25.10.4) and packages each step as a versioned Docker, Singularity, or Apptainer container image.
\texttt{DIA-NN}~1.8.1 is distributed publicly via BioContainers; because the \texttt{DIA-NN} license restricts redistribution of later releases, containers for \texttt{DIA-NN}~1.9 and above are built locally from the recipes in \url{https://github.com/bigbio/quantms-containers} (see Experimental Procedures).
Pipeline version v2.1.0 (release codename ``Sao Pablo'', 2026-05-08) was used for the analyses reported in this paper.
Analysis scripts that generated the figures in this manuscript are available at \url{https://github.com/bigbio/quantmsdiann-paper} (this repository).

\section*{Supplemental Data}
This article contains supplemental data.
Additional file~1 provides Table~S1 (the full PRIDE accession list), per-process runtime and resource envelopes (Figs.~S1--S2), ProteoBench supplementary panels (Figs.~S3--S4), and per-cohort completeness, peptide-per-protein and gene-versus-accession overlap figures (Figs.~S5--S13).

\section*{Conflict of Interest}
The authors declare no competing interests.

\section*{Acknowledgments}
\placeholder{We thank the Brunner, Derks, Leduc, Mund, Wang, and Coscia teams for depositing their data publicly. We thank the nf-core community and the OpenMS and DIA-NN developer teams for their ongoing work.}
M.B.\ is supported by the Science and Technology Innovation Key R\&D Program of Chongqing (CSTB2023TIAD-STX0002).
\placeholder{Funding sources and grant numbers for the remaining authors.}

\section*{Author Contributions}
QXY, YS and CD developed the pipeline.
HW helped build the containers and contributed to the workflow implementation and manuscript writing.
AL annotated the SDRF metadata and helped run the workflow.
TS, FC, NS and JN assisted with the DIA-NN reanalyses and manuscript writing; TS additionally contributed to the workflow design, and JN tested the tool and identified multiple issues.
OK provided feedback on the plexDIA datasets.
VD provided DIA-NN expertise and guidance on cross-version reproducibility.
MB contributed to manuscript writing and supervised the students; MB and YPR conceived and supervised the project.
QXY and YPR wrote the original draft, with input from all authors.
All authors read and approved the final manuscript.

\section*{Abbreviations}
\begin{description}[leftmargin=5.5em,style=nextline,itemsep=0pt,parsep=0pt]
\item[CLI] command-line interface
\item[DDA] data-dependent acquisition
\item[DIA] data-independent acquisition
\item[DIA-NN] data-independent acquisition neural networks (DIA search and quantification software)
\item[diaPASEF] data-independent acquisition with parallel accumulation--serial fragmentation
\item[FDR] false discovery rate
\item[HPC] high-performance computing
\item[HYE] human, yeast, and \textit{E.\ coli} (ProteoBench three-proteome mixture)
\item[IQR] interquartile range
\item[LSF] Load Sharing Facility (HPC scheduler)
\item[MS] mass spectrometry
\item[MS1/MS2] first-/second-stage mass spectrometry (precursor/fragment scans)
\item[PASEF] parallel accumulation--serial fragmentation
\item[QPX] Quantitative Proteomics eXchange
\item[SDRF] Sample and Data Relationship Format
\end{description}
