\quantmsdiann{} automates the path from a deposited DIA dataset to a reproducible quantification table, at a scale that supports systematic reanalysis of public DIA archives.
The SDRF-first design encourages experiment annotation upstream of analysis and produces metadata that is interoperable with the wider proteomics ecosystem~\cite{dai2021sdrf}; the \qpx{} archive bundles the SDRF, harmonised quantification tables and pipeline provenance into a single Parquet container so that reanalysed results can be archived (e.g.\ on Zenodo) and re-queried alongside the original raw data.
By focusing on \texttt{DIA-NN} rather than the multi-engine scope of the related bigbio/quantms pipeline~\cite{dai2024quantms}, \quantmsdiann{} achieves a leaner dependency surface and faster pipeline-level execution for the DIA-only regime, where the dominant cost is per-file \texttt{DIA-NN} invocation rather than across-engine comparison.

This parallel structure also shapes how the workflow scales. End-to-end wall-clock is the sum of a largely fixed serial overhead (chiefly in-silico spectral-library generation from the FASTA, the single largest non-parallel step, together with cohort-level consolidation) and the per-file analysis stages, which run as independent tasks across the cluster (Fig.~\ref{fig:architecture}b) and whose elapsed time grows only sub-linearly with cohort size as files are processed in parallel waves (per-step envelopes in Additional file~1: Fig.~S1). Time-to-finish is therefore dominated by the fixed overhead for small cohorts and stays within a few hours even for the largest deposits, so a small dataset is not necessarily faster than a much larger one (Fig.~\ref{fig:architecture}c); the serial overhead itself scales with the predicted-library size, growing for wide variable-modification searches such as phosphoproteomics, rather than with the number of files.

The demonstrations stress the workflow along three orthogonal axes: analytical fidelity (ProteoBench; Fig.~\ref{fig:proteobench}), recovery against published per-study quantifications (independent reanalyses; Fig.~\ref{fig:cell-line-reanalysis}), and archive-scale throughput (pan-cohort; Fig.~\ref{fig:pancohort}).
The recovery results suggest that the value of DIA reanalysis is real but time-limited (largest for deposits originally analysed with older or non-\texttt{DIA-NN} engines and shrinking toward parity for recently processed cohorts), a regime an SDRF-driven workflow is suited to address as new \texttt{DIA-NN} versions ship.
The workflow already supports broader single-cell and laser-microdissected spatial acquisition modes (Fig.~\ref{fig:architecture}a, library-handling branch), and SDRFs for the Brunner~\cite{brunner2022}, Derks~\cite{derks2023}, Leduc~\cite{leduc2024}, Wang~\cite{wang2025}, Mund~\cite{mund2022} and Makhmut~\cite{makhmut2023} datasets are in preparation.

The cross-\texttt{DIA-NN}-version reproducibility view enabled by \quantmsdiann{} is, to our knowledge, the first pipeline-level treatment of this question for \texttt{DIA-NN}.
We expect this to become increasingly important as \texttt{DIA-NN} evolves: a deposited single-cell DIA dataset analysed with \diannversion{1.8} today and re-analysed with \diannversion{2.x} in two years should not produce different biological conclusions for the same data and parameters. In current practice such re-analyses are rarely performed, because the bespoke per-study scripts make them prohibitively expensive, a barrier the single-invocation audit of Section~\ref{sec:proteobench} removes.

\quantmsdiann{} has several limitations.
First, \quantmsdiann{} does not currently include batch correction, cell-type assignment, or differential-abundance modules; these are best handled in downstream R or Python packages such as scp~\cite{vanderaa2023}, which can read the \qpx{} or MSstats outputs directly.
Second, library generation for plexDIA channel deconvolution remains internal to \texttt{DIA-NN}, and \quantmsdiann{} inherits any limitations of that step.
Third, the \qpx{} format will continue to evolve as community feedback accumulates; backward compatibility is maintained through schema versioning.
Fourth, beyond the single-cell oocyte plexDIA reanalysis presented here, broader single-cell and laser-microdissected spatial DIA demonstrations (Fig.~\ref{fig:architecture}a, library-handling branch) are not evaluated in depth in this work.

Finally, we note that \quantmsdiann{} is complementary to, rather than competing with, existing single-cell DIA workflows.
The benchmarking work of Wang \textit{et al.}~\cite{wang2025} compared multiple search engines and rescoring strategies; \quantmsdiann{} adopts the broadly competitive \texttt{DIA-NN} engine and focuses on the orthogonal problem of orchestrating that engine reproducibly at scale.
Similarly, earlier nf-core DIA efforts such as DIAproteomics~\cite{bichmann2021diaproteomics} and, more recently, the MHCquant2 pipeline~\cite{scheid2025mhcquant2} address analogous orchestration problems for DIA and immunopeptidomics.
Together, these contributions extend the nf-core proteomics ecosystem to cover the major MS-based proteomics modalities behind a common SDRF-driven interface.
