Data-independent acquisition (DIA) is now the dominant mass-spectrometry strategy for proteomics at scale, and \texttt{DIA-NN}~\cite{demichev2020} is one of the most widely adopted analysis engine across timsTOF~\cite{brunner2022,meier2020}, Astral, and Orbitrap platforms.
Recent benchmarks confirm \texttt{DIA-NN}'s competitiveness for single-cell DIA workflows~\cite{wang2025}, and versions 1.9.x and 2.x have expanded support for plexDIA, timsTOF parallel accumulation-serial fragmentation (PASEF), and vendor-native raw-file reading.
Despite this analytical maturity, the majorty of DIA-NN users use the tool in Desktop applications; which limited the scalability of the tool for large-scale datasets. In addition; deployments in HPC and cloud infrastructures remain artisanal using shell scripts, ad-hoc directory conventions, and hand-curated sample sheets, so the same raw data analysed by two laboratories, or with two \texttt{DIA-NN} releases, can yield non-overlapping quantification tables that are difficult to reconcile.

Beyond per-study analysis, the value of public proteomics archives such as PRIDE Archive~\cite{deutsch2026proteomexchange} lies as much in systematic reanalysis as in original deposition.
In 2024 we published bigbio/quantms~\cite{dai2024quantms}, a Nextflow pipeline that standardised multi-engine DIA and DDA analyses behind a Sample and Data Relationship Format (SDRF)~\cite{dai2021sdrf} input on the nf-core framework~\cite{ewels2020nfcore}. quantms was, however, built around \texttt{DIA-NN}~1.8.1 with a fixed parameter set, and its DDA/DIA breadth made extending DIA-specific functionality difficult: new \texttt{DIA-NN} parameters conflicted with DDA options, versions were hard-coded, sample-level acquisition parameters were not properly propagated to per-run \texttt{DIA-NN} invocations, and raw-file conversion was mandatory. Iterative reanalysis at archive scale requires a workflow that combines the reproducibility of nf-core, the sensitivity of current \texttt{DIA-NN} releases, the metadata discipline of SDRF, and a parallelisation strategy capable of processing thousands of raw files.

Here, we present \quantmsdiann{}, an independent SDRF-driven Nextflow pipeline dedicated to \texttt{DIA-NN} that emerged from the need to extend DIA-specific features beyond what the original bigbio/quantms pipeline could accommodate. While \quantmsdiann{} shares its SDRF-first design philosophy and parts of the configuration-generation tooling with quantms, the workflow, container set, and module library are developed and released independently: it ships container-pinned profiles for all DIANN versions, selectable at runtime and resolved using Docker or Singularity depending on the enviroment; whitouh the need of dependency instalations. \quantmsdiann{} natively supports all vendor formats, and when conversion is needed (e.g. Sciex) the workflow handles it seamlessly with the conversion to mzML. The parallelisation design allows a 2{,}300-run single-cell dataset to be reanalysed in 2.4~hours on an HPC cluster of 300 nodes. Downstream, \quantmsdiann{} exports a harmonised \qpx{} (Quantitative Proteomics eXchange) Parquet-based archive that bundles SDRF, identifications, quantifications, and pipeline provenance into a single queryable artefact for downstream analysis and long-term archival alongside the standard \texttt{DIA-NN} reports and MSstats input.
We demonstrate the pipeline along three axes: ProteoBench~\cite{devreese2025proteobench} benchmarks across four DIA modalities and five \texttt{DIA-NN} releases; three independent bulk-DIA cell-line reanalyses (PXD003539, PXD004701, PXD030304); and a pan-cohort of five public DIA cell-line reanalyses integrated under a single SDRF-driven invocation.
